Capstone Movielens Report

### 2025-11-28 reports/capstone-movielens.site/02.5.UME-model.Rmd
`UMGE Model Regularization: Fine-tuning` section described.

::: {.noteblock data-latex=""}
The functions [tune.model_param](#func.tune.model_param) and [regularize.test_lambda.UM_effect.cv](#func.regularize.test_lambda.UM_effect.cv) are described in the sections *Common Helper Functions /* [ Regularization Functions](#appndx_a.CHF.regularization) and  *Models Training: Support Functions /* [UME Model: Regularization] of [Appendix A: Support Functions], respectively.
:::

@@

put_log1("Preliminary regularization set-up of `lambda`s range for the UME Model has been completed
for the %1-Fold Cross-Validation samples.",
CVFolds_N)

```
```{r}
str(cv.UME.preset.result)
```

>>>>

::: {.noteblock data-latex=""}
The functions [tune.model_param](#func.tune.model_param) and [regularize.test_lambda.UM_effect.cv](#func.regularize.test_lambda.UM_effect.cv) are described in the sections *Common Helper Functions /* [Regularization Functions](#appndx_a.CHF.regularization) and  *Models Training: Support Functions /* [UME Model: Regularization] of [Appendix A: Support Functions], respectively.
:::

@@

put_log1("Preliminary regularization set-up of `lambda`s range for the UME Model has been completed
for the %1-Fold Cross-Validation samples.",
CVFolds_N)
```
```{r}
str(cv.UME.preset.result)
cv.UME.preset.result$best_result
```

### 2025-11-28 reports/capstone-movielens.site/02.6.UMGE-model.Rmd
`UMGE Model Regularization: Fine-tuning` section described.

The plot below shows strong evidence of a genre effect (for illustrative purposes, the plot shows only categories with more than 20, 000 ratings).

@@

::: {.noteblock data-latex=""}
The complete source code of builing and training the current model is available in the [Model building: User+Movie+Genre Effect](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L1149) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R) script on _GitHub_.
:::

Below, we provide the most significant part of the code for training our model using the `5-Fold Cross Validation` method:
```{r, fn.train_user_movie_genre_effect.cv,   eval=FALSE}
cv.UMG_effect <- train_user_movie_genre_effect.cv()
```
```{r, hist.cv.UMG_effect }
str(cv.UMG_effect)

@@

We can now construct predictors and calculate the _RMSE_ of the current model using the
[calc_user_movie_genre_effect_RMSE.cv](#func.calc_user_movie_genre_effect_RMSE.cv) function described above:

```{r, eval=FALSE}
cv.UMG_effect.RMSE <- calc_user_movie_genre_effect_RMSE.cv(cv.UMG_effect)

@@

@@

As stated in the [UME Model Regularization] section, we implement the _Regularization_ method for our models in the following three steps:

  1. **Pre-configuration:** Preliminary determination of the optimal range of $\lambda$ values for the `5-Fold Cross Validation` samples;

  2. **Fine-tuning:** figuring out the value of $\lambda$ that minimizes the model's RMSE.

  3. **Retraining:** retraining the model with the best value of the parameter $\lambda$ obtained in the previous step.

#### UMGE Model Regularization: Support Functions
\

::: {.noteblock data-latex=""}
The [regularize.test_lambda.UMG_effect.cv](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UMG-effect.functions.R#L146) function described below are defined in the [Regularization](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UMG-effect.functions.R#L145) section of the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UMG-effect.functions.R#L146) script.
:::

##### [regularize.test_lambda.UMG_effect.cv](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UMG-effect.functions.R#L146) Function {#func.regularize.test_lambda.UMG_effect.cv}
\

This function calculates _RMSE_ of the _UMGE Model_ using _5-Fold Cross Validation_ method for the given $\lambda$ parameter value:
```{r, eval=FALSE}
regularize.test_lambda.UMG_effect.cv <- function(lambda){
  if (is.na(lambda)) {
    stop("Function: regularize.test_lambda.user_movie_genre_effect.cv
`lambda` is `NA`")
  }

  umg_effect <- train_user_movie_genre_effect.cv(lambda)
  calc_user_movie_genre_effect_RMSE.cv(umg_effect)
}
```

::: {.noteblock data-latex=""}
Note that we reuse the function [train_user_movie_genre_effect.cv](#func.train_user_movie_genre_effect.cv) calling it from the [regularize.test_lambda.UMG_effect.cv](#func.regularize.test_lambda.UMG_effect.cv), but now with the $\lambda$ parameter different from the default ('lambda = 0') value.
:::

@@

@@

Below we provide the most significant part of the code that performs this operation:
```{r, eval=FALSE}
  put_log1("Preliminary setting-up of `lambda`s range for %1-Fold Cross Validation samples...",
           CVFolds_N)

  start <- put_start_date()
  lambdas <- seq(0, 0.2, 0.01)
  cv.UMGE.preset.result <- 
    tune.model_param(lambdas, regularize.test_lambda.UMG_effect.cv)
  put_end_date(start)
  put_log1("Preliminary regularization set-up of `lambda`s range for the UMGE Model has been completed
for the %1-Fold Cross Validation samples.",
CVFolds_N)
```
```{r}
str(cv.UMGE.preset.result)
cv.UMGE.preset.result$best_result
```

::: {.noteblock data-latex=""}
The complete version of the source code provided in this section can be found in the [UMGE Model Regularization: Pre-configuration](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L814) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R) script.
:::

@@

::: {.noteblock data-latex=""}
We use the custom data visualization function [data.plot](#func.data.plot) described in the [Data Helper Functions](#appndx_a.data_helper.functions) section of [Appendix A: Support Functions], which is defined in the [Data Visualization](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L447) section of the [data.helper.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L592) script on _GitHub_.
:::

@@

@@

@@

UMG_effect.loop_starter <- c(endpoints["start"], 
                            endpoints["end"], 
                            8)
UMG_effect.loop_starter
#> [1] 0.0   0.1   8.0

UMGE.rglr.fine_tune.cache.base_name <- "UMGE.rglr.fine-tuning"

@@

UMGE.rglr.fine_tune.RMSE.best <- UMGE.rglr.fine_tune.results$best_result["best_RMSE"]
# best_RMSE 
#  0.872973

```
```{r}
str(UMGE.rglr.fine_tune.results)
UMGE.rglr.fine_tune.results$best_result
```

::: {.noteblock data-latex=""}
The complete version of the source code provided in this section can be also found in the [Fine-tuning Step of the Regularization Method for the User+Movie Model](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L867) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R) script on _GitHub_.
:::

>>>>

The plot below shows strong evidence of a genre effect (for illustrative purposes, the plot shows only categories with more than 40, 000 ratings).

@@

::: {.noteblock data-latex=""}
The complete source code of builing and training the current model is available in the [UMGE Model Building](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L878) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L878) script on _GitHub_.
:::

Below, we provide the most significant part of the code for training our model using the `5-Fold Cross Validation` method:
```{r, fn.train_user_movie_genre_effect.cv,   eval=FALSE}
cv.UMG_effect <- train_user_movie_genre_effect.cv()
```

::: {.noteblock data-latex=""}
In the code snippet above we use the [train_user_movie_genre_effect.cv](#func.train_user_movie_genre_effect.cv) function described in the *Models Training: Support Functions /* [UMGE Model: Utility Functions] section of [Appendix A: Support Functions].
:::

```{r, hist.cv.UMG_effect }
str(cv.UMG_effect)

@@

We can now construct predictors and calculate the _RMSE_ of the *UMGE Model* using the
following code:
```{r, eval=FALSE}
cv.UMG_effect.RMSE <- calc_user_movie_genre_effect_RMSE.cv(cv.UMG_effect)

@@

::: {.noteblock data-latex=""}
In the code snippet above, we use the following functions described in [Appendix A: Support Functions]:

  - [calc_user_movie_genre_effect_RMSE.cv](#func.calc_user_movie_genre_effect_RMSE.cv) (described in the *Models Training: Support Functions /* [UMGE Model: Utility Functions] section) to calculate the *RMSE* for the *UMGE Model*;

  - [RMSEs.AddRow](#func.RMSEs.AddRow) and [RMSE_kable](#func.RMSE_kable) (described in the *Common Helper Functions /* [Result RMSEs Tibble Functions] section) to update and print the table above.
:::

@@

As we decided earlier (and outlined in the [UME Model Regularization] section), we implement the *Regularization method* for our models in the following three steps:

  1. **Pre-configuration:** Preliminary determination of the optimal range of *regularization parameter* $\lambda$ values for the *K-Fold Cross-Validation* samples;

  2. **Fine-tuning:** Figuring out the best value of $\lambda$ with the highest possible accuracy that minimizes the *RMSE* of the model.

  3. **Retraining:** Retraining the model with the best value of $\lambda$ determined in the previous step.

@@

::: {.noteblock data-latex=""}
The complete version of the source code provided in this section can be found in the [UMGE Model Regularization: Pre-configuration](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L961) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L961) script.
:::

As we explain earlier in the [UME Model Regularization: *Pre-configuration*], we need to do some pre-configuration to determine a suitable range of $\lambda$ for subsequent fine-tuning of our current model.

Below we provide the most significant part of the code that performs this operation:

::: {.noteblock data-latex=""}
The functions [tune.model_param](#func.tune.model_param) and [regularize.test_lambda.UMG_effect.cv](#func.regularize.test_lambda.UMG_effect.cv) are described in the sections *Common Helper Functions /* [ Regularization Functions](#appndx_a.CHF.regularization) and  *Models Training: Support Functions /* [UMGE Model: Regularization] of [Appendix A: Support Functions], respectively.
:::

@@

Below we provide the most significant part of the code that performs this operation:
```{r, eval=FALSE}
lambdas <- seq(0, 0.2, 0.01)

cv.UMGE.preset.result <- 
  tune.model_param(lambdas, regularize.test_lambda.UMG_effect.cv)

put_log1("Preliminary regularization set-up of `lambda`s range for the UMGE Model has been completed
for the %1-Fold Cross Validation samples.",
CVFolds_N)
```
```{r}
str(cv.UMGE.preset.result)
cv.UMGE.preset.result$best_result
```

::: {.noteblock data-latex=""}
The custom functions [tune.model_param](#func.tune.model_param) and [regularize.test_lambda.UMG_effect.cv](#func.regularize.test_lambda.UMG_effect.cv) used in the code snippet above are described in the sections *Common Helper Functions /* [Regularization Functions](#appndx_a.CHF.regularization) and  *Models Training: Support Functions /* [UMGE Model: Regularization] of [Appendix A: Support Functions], respectively.
:::

@@

::: {.noteblock data-latex=""}
In the code snippet above, we use the custom data visualization function [data.plot](#func.data.plot) described in the *Data Helper Functions /* [Data Visualization Functions] section of [Appendix A: Support Functions].
:::

@@

::: {.noteblock data-latex=""}
The complete version of the source code provided in this section can be found in the [UMGE Model Regularization: Fine-tuning](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L1019) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L1019) script on _GitHub_.
:::

@@

We are now ready to perform the fine-tuning step of our model _regularization_ process to determine the best value for the $\lambda$ parameter.
For this purpose, we will use the helper function [model.tune.param_range](#func.model.tune.param_range), passing the [regularize.test_lambda.UM_effect.cv](#func.regularize.test_lambda.UM_effect.cv) function (already mentioned above in the [UME Model Regularization: *Pre-configuration*] section) as the value of its argument [fn_tune.test.param_value](#func.model.tune.param_range.args.fn_tune.test.param_value).

@@

UMG_effect.loop_starter <- c(endpoints["start"], 
                            endpoints["end"], 
                            8)
```

::: {.noteblock data-latex=""}
The helper function [get_fine_tune.param.endpoints](#func.get_fine_tune.param.endpoints) used in code snippet above is described in the sections *Common Helper Functions /* [ Regularization Functions](#appndx_a.CHF.regularization) of [Appendix A: Support Functions].
:::

```{r}
UMG_effect.loop_starter
UMGE.rglr.fine_tune.cache.path
```
```{r, eval=FALSE}
UMGE.rglr.fine_tune.cache.base_name <- "UMGE.rglr.fine-tuning"

@@

UMGE.rglr.fine_tune.RMSE.best <- UMGE.rglr.fine_tune.results$best_result["best_RMSE"]
```

::: {.noteblock data-latex=""}
The custom functions [model.tune.param_range](#func.model.tune.param_range) and [regularize.test_lambda.UM_effect.cv](#func.regularize.test_lambda.UM_effect.cv) used in the code snippet above are described in the sections *Common Helper Functions /* [Regularization Functions](#appndx_a.CHF.regularization) and  *Models Training: Support Functions /* [UME Model: Regularization] of [Appendix A: Support Functions], respectively.
:::

```{r, echo=FALSE}
writeLines("*** Fine-tuning results object data structure ***")
```
```{r}
str(UMGE.rglr.fine_tune.results)
```
```{r, echo=FALSE}
writeLines("*** Fine-tuning: best results ***")
```
```{r}
UMGE.rglr.fine_tune.results$best_result
UMGE.rglr.fine_tune.RMSE.best
```
